---
title: "A collection of data on the Auckland Volcanic Field"
author: "AHC"
date: "2023-11-05"
output: 
  html_document:
    code_folding: show
license: "GPL-3.0-or-later"
editor_options: 
  markdown: 
    wrap: 72
---

## Background

```{r packages, warning=F, message=F}
# install.packages(c("tidyverse", "httr", "XML"))
library(tidyverse)
  # `tidyverse` for data cleaning, usability, and output
library(httr)
library(XML)
  # `httr` and `XML` for data import
```

This is whole thing is drawn from the [Wikipedia article on the Auckland
Volcanic
Field](https://en.m.wikipedia.org/w/index.php?title=Auckland_volcanic_field&oldid=1146734511).
I'm using this as practice in extracting data from the web, making it
pretty and presentable, all while having a semi-practical use to import
geospatial specific points for all of the volcanoes. (Anti-skill loss
project my beloved \<3).

```{r import}
# Import data from Wikipedia
# From user:schnee response on https://stackoverflow.com/questions/7407735/importing-wikipedia-tables-in-r

url <- "https://en.m.wikipedia.org/w/index.php?title=Auckland_volcanic_field&oldid=1146734511"
  # This is a permanent link to the exact data I used (accessed on 2023-07-17), use the following at your own risk:
  # url <- "https://en.m.wikipedia.org/wiki/Auckland_volcanic_field"

r <- GET(url)

doc <- readHTMLTable(
  doc=content(r, "text"), header = T)
```

Now that we have imported the data, let's clean it up a little bit.

```{r cleaning, warning=F}
volcanoes <- 
  tibble(doc[[2]]) %>% 
  
  # Cleaning headers -----------------------------------------------------------
  rename(
    Volcanoes = `Volcanoes\n`,
    Age = `Age (thousand years)[38]`,
    Height = `Height\n`,
    Location = `Location (Coordinates)\n`,
    Refs = `Refs\n`,
    Images = `Images\n`
  ) %>% 
  
  # Feel free to keep the Refs & Images, I don't find them useful --------------
  select(!c(Refs, Images)) %>% 
  
  # Cleaning Height data -------------------------------------------------------
  separate(col = Height, into = c("Height (m)", NA, "Height (ft)", NA, "Quarried")) %>% 
  mutate(
    `Height (m)` = as.integer(`Height (m)`),
    `Height (ft)` = as.integer(`Height (ft)`),
    Quarried = ifelse(Quarried == "quarried", "T", "F"),
    Quarried = replace_na(Quarried, "F"),
    Quarried = as.logical(Quarried)
      # I don't know if multiple mutates in a row is kosher, but it works for me
  ) %>% 
  
  # Cleaning coordinates, only need one of these after all ---------------------
  separate(col = Location, into = c(NA, NA, "Location"), sep = "/") %>% 
    
  # Cleaning age ---------------------------------------------------------------
  separate(col = Age, into = c("Age", "Uncertainty (±)"), sep = "±")
  
  # Manually update the values bc it's simpler than finding a particular function to do it
  volcanoes[53,2] <- "30.55"
  volcanoes[53,3] <- "0.45"
  volcanoes[3,2] <- "130"
  volcanoes[27,2] <- "45"
  volcanoes[31,2] <- "130"
  volcanoes[33,2] <- "0.62"
  volcanoes[47,2] <- "28.0"
    # Yes, this is horribly messy and gross. How'd you know? (but it works soooooo idc)

  volcanoes <- volcanoes %>% 
  mutate(
    Age = as.numeric(Age),
    Age = Age * 1000,
    Age = as.integer(Age),
    `Uncertainty (±)` = as.numeric(`Uncertainty (±)`),
    `Uncertainty (±)` = `Uncertainty (±)` * 1000,
    `Uncertainty (±)` = as.integer(`Uncertainty (±)`)
  )
  
  # That one funny name that wouldn't parse perfectly
  volcanoes[52,1] <- "Whakamuhu / Saint Heliers / Glover Park"

# Data output for however you want it ------------------------------------------
write_csv(volcanoes, "volcanoes.csv")
  
# Print dataset ================================================================

knitr::kable(volcanoes)
```

## Now what?

Well, ain't that the question. For me, I created this because I plan on
visiting all `r nrow(volcanoes)`(!) historic volcanoes in Auckland.
Seems fun, and I don't have anything better to do. I went through the
hassle of making this into a gmaps list, but it's deeply inaccurate (I
attached each point to a nearby place rather than its coordinates) so I
used the code below in [Volcanoes to visit generator] to add it to a
Google My Maps. However, the points seem a little inaccurate... I will
report back from my adventures with any important changes.

Feel free to play with this, if you want a neat little dataset or some
practice in R by trying to reproduce this.

Some fun stats (now easy to do with the data already in R):

```{r stats, echo=FALSE}
volcanoes %>% 
  summarize(
    `Number of volcanoes` = n(),
    `Total Height (m)` = sum(`Height (m)`, na.rm = TRUE),
    `Total Height (ft)` = sum(`Height (ft)`, na.rm = TRUE),
    `Mean Age` = mean(Age, na.rm = TRUE),
    `Number of Quarried sites` = sum(Quarried, na.rm = TRUE)
  ) %>% 
  mutate(
    `Mean Age` = as.integer(`Mean Age` / 1000)
  ) %>% 
  rename(
    `Mean Age (thousand years)` = `Mean Age`
  ) %>% 
  knitr::kable()
```

Anyways. Have fun, enjoy the data! I might leave an update on any
eventual updates, or perhaps a little table that is updated with info on
the sites I have visited so far.

Peace and love \<3 - alice

Version 1.1: added volcano tracker.

## Volcano Tracker

```{r visited, message=F}
# Yes I tried to make a Shiny app. Yes I realized about an hour into it that it would not be static.
# Yes this hurts me.
volcanoes <- read_csv("volcanoes.csv")
visited <- tibble(
  Volcanoes = c("Maungauika / North Head", "Takarunga / Mount Victoria", "Albert Park Volcano", "Maungakiekie / One Tree Hill", "Te Kōpuke / Tītīkōpuke / Mount St John", "Ōhinerau / Mount Hobson", "Ohuiarangi / Pigeon Mountain", "Te Kopua-o-Matakamokamo / Tank Farm / Tuff Crater", "Onepoto", "Maungarei / Mount Wellington", "Taurere / Taylors Hill", "Whakamuhu / Saint Heliers / Glover Park", "Maungawhau / Mount Eden", "Takaroro / Mount Cambria", "Te Pou Hawaiki", "Te Tātua-a-Riukiuta / Three Kings", "Te Hopua-a-Rangi / Gloucester Park", "Te Pane-o-Mataaho / Māngere Mountain", "Māngere Lagoon", "Ōrākei Basin", "Maungarahiri / Little Rangitoto", "Pukekawa / Auckland Domain", "Grafton Volcano", "Ash Hill", "Wiri Mountain / Matukutūruru", "Matukutureia / McLaughlins Mountain", "Puhinui Craters"),
  Visited = T, # Redundant column for future filters or whatever idk
  `Date visited` = c("2023-07-11", "2023-07-11", "2023-07-17", "2023-07-22", "2023-07-22", "2023-07-22", "2023-07-30", "2023-08-03", "2023-08-03", "2023-08-27", "2023-08-27", "2023-08-27", "2023-09-01", "2023-09-17", "2023-09-27", "2023-10-13", "2023-10-14", "2023-10-14", "2023-10-14", "2023-11-04", "2023-11-04", "2023-10-26", "2023-08-05", "2023-11-06", "2023-11-06", "2023-11-06", "2023-11-06")
)

bigvisited <- inner_join(volcanoes, visited) %>% 
  select(Volcanoes, `Date visited`, Age, `Height (ft)`, Location) %>% 
  mutate(
    `Date visited` = as.Date(`Date visited`)
  )

summaryvisited <- inner_join(volcanoes, visited) %>% 
  summarize(
    Location = paste0("n = ", n(), " / ", nrow(volcanoes), " (", round((n() / nrow(volcanoes)) * 100), "%)"),
    `Height (ft)` = sum(`Height (ft)`, na.rm = T),
    Age = as.integer(mean(Age, na.rm = T)),
    Volcanoes = "Mean Age / Total Height"
  )

knitr::kable(
  bind_rows(bigvisited, summaryvisited) %>% 
    arrange(`Date visited`, Volcanoes)
)
```

Tracker last updated 2023-11-06.

Version 1.3 added `to visit` csv generator for maps.

### Volcanoes to visit generator

You know how I mentioned wanting to have a nice little export into
google or something to easily know which volcanoes I still had to visit?
Now is the time!

```{r to visit, message = F}
volcanoes <- volcanoes %>% 
  mutate(
    Lat = str_split_i(Location, ";", 1),
    Long = str_split_i(Location, ";", 2)
  )

left_join(volcanoes, visited) %>% 
  filter(is.na(Visited)) %>% 
  write_csv(., "tovisit.csv")
```

And, now that we have that with Lat and Long specified in columns, we
can import it into a [Google My
Map](https://www.google.com/maps/d/u/0/edit?mid=1IJyYLAL8ikhqOcV66Ka2mTrnSG_Z_qk&usp=sharing)
(because you can't with normal gmaps?). A neat thing I found is that you
can re-import specific layers, which is nice once a couple of volcanoes
are ticked off the list.

## Code availability

The source code for this project is available at
<https://github.com/legallyahc/volcanic>, with all work done right in
this RMarkdown file (`dataset.rmd`) :3.

## Copyright

This project is based on information from Wikipedia which was authored
by Wikipedia contributors and is licensed under [CC-BY-SA
4.0](https://creativecommons.org/licenses/by-sa/4.0/). This project is
licensed under the [GNU General Public License
v3](https://www.gnu.org/licenses/gpl-3.0.html) or later, a compatible
license with CC-BY-SA 4.0.

<br>

Originally published on 2023-07-16. Version 1.3.1.

<!-- This is included for publishing on my website, feel free to exclude -->

##### [Back to What?](../what.html)
