---
title: "A collection of data on the Auckland Volcanic Field"
author: "AHC"
date: "2023-08-01"
output: 
  html_document:
    code_folding: show
license: "GPL-3.0-or-later"
---

## Packages

```{r packages, warning=F, message=F}
#install.packages(c("tidyverse", "httr", "XML"))
# For data cleaning, usability, and output
library(tidyverse)

# To import data from Wikipedia
library(httr)
library(XML)
```

## Background

This is whole thing is drawn from the [Wikipedia article on the Auckland Volcanic Field](https://en.m.wikipedia.org/w/index.php?title=Auckland_volcanic_field&oldid=1146734511). 
I'm using this as practice in extracting data, making it pretty and presentable, 
and hopefully being able to use this practically as well to import nice reference 
points of all of the volcanoes. (Anti-skill loss project my beloved <3).

```{r import}
# Import data from Wikipedia
# From user:schnee response on https://stackoverflow.com/questions/7407735/importing-wikipedia-tables-in-r

url <- "https://en.m.wikipedia.org/w/index.php?title=Auckland_volcanic_field&oldid=1146734511"
        # This is a permanent link to the exact data I used (accessed on 2023-07-17), use the following at your own risk:
#url <- "https://en.m.wikipedia.org/wiki/Auckland_volcanic_field"

r <- GET(url)

doc <- readHTMLTable(
  doc=content(r, "text"), header = T)
```

Now that we have imported the data, let's clean it up a little bit.

```{r cleaning, warning=F}
volcanoes <- 
  tibble(doc[[2]]) %>% 
  
  # Cleaning headers -----------------------------------------------------------
  rename(
    Volcanoes = `Volcanoes\n`,
    Age = `Age (thousand years)[38]`,
    Height = `Height\n`,
    Location = `Location (Coordinates)\n`,
    Refs = `Refs\n`,
    Images = `Images\n`
  ) %>% 
  
  # Feel free to keep the Refs & Images, I don't find them useful --------------
  select(!c(Refs, Images)) %>% 
  
  # Cleaning Height data -------------------------------------------------------
  separate(col = Height, into = c("Height (m)", NA, "Height (ft)", NA, "Quarried")) %>% 
  mutate(
    `Height (m)` = as.integer(`Height (m)`),
    `Height (ft)` = as.integer(`Height (ft)`),
    Quarried = ifelse(Quarried == "quarried", "T", "F"),
    Quarried = replace_na(Quarried, "F"),
    Quarried = as.logical(Quarried)
      # I don't know if multiple mutates in a row is kosher, but it works for me
  ) %>% 
  
  # Cleaning coordinates, only need one of these after all ---------------------
  separate(col = Location, into = c(NA, NA, "Location"), sep = "/") %>% 
    
  # Cleaning age ---------------------------------------------------------------
  separate(col = Age, into = c("Age", "Uncertainty (±)"), sep = "±")
  
  # Manually update the values bc u can't be fucked to figure out the functions to do it
  volcanoes[53,2] <- "30.55"
  volcanoes[53,3] <- "0.45"
  volcanoes[3,2] <- "130"
  volcanoes[27,2] <- "45"
  volcanoes[31,2] <- "130"
  volcanoes[33,2] <- "0.62"
  volcanoes[47,2] <- "28.0"
    # Yes, this is horribly messy and gross. How'd you know? (but it works soooooo idc)

  volcanoes <- volcanoes %>% 
  mutate(
    Age = as.numeric(Age),
    Age = Age * 1000,
    Age = as.integer(Age),
    `Uncertainty (±)` = as.numeric(`Uncertainty (±)`),
    `Uncertainty (±)` = `Uncertainty (±)` * 1000,
    `Uncertainty (±)` = as.integer(`Uncertainty (±)`)
  )
  
  # That one funny name
  volcanoes[52,1] <- "Whakamuhu / Saint Heliers / Glover Park"

# Data output for however you want it ------------------------------------------
write_csv(volcanoes, "volcanoes.csv")
  
# Print dataset ================================================================

knitr::kable(volcanoes)
```

## Now what?

Well, ain't that the question. For me, I created this because I plan on visiting all `r nrow(volcanoes)`(!) 
historic volcanoes in Auckland. Seems fun, and I don't have anything better to do. I went through 
the hassle of making this into a google list, but it's deeply inaccurate (I attached each point to a nearby
place rather than its coordinates) so I plan on remaking based on the coordinates provided in the Wiki page,
and I might try to find some fun OSM (Open Street Map) based thing. However, the points seem a little 
inaccurate... I will report back from my adventures with any important changes.

Feel free to play with this, if you want a neat little dataset or some practice in R by trying to reproduce 
this.

Some fun stats (now easy to do with the data already in R):
```{r stats, echo=FALSE}
volcanoes %>% 
  summarize(
    `Number of volcanoes` = n(),
    `Total Height (m)` = sum(`Height (m)`, na.rm = TRUE),
    `Total Height (ft)` = sum(`Height (ft)`, na.rm = TRUE),
    `Mean Age` = mean(Age, na.rm = TRUE),
    `Number of Quarried sites` = sum(Quarried, na.rm = TRUE)
  ) %>% 
  mutate(
    `Mean Age` = as.integer(`Mean Age` / 1000)
  ) %>% 
  rename(
    `Mean Age (thousand years)` = `Mean Age`
  ) %>% 
  knitr::kable()
```
Anyways. Have fun, enjoy the data! I might leave an update on any eventual updates,
or perhaps a little table that is updated with info on the sites I have visited so far. 

Peace and love <3 - alice

Version 1.1: added volcanoe tracker.

## Volcanoe Tracker

```{r visited, message=F}
# Yes I tried to make a Shiny app. Yes I realized about an hour into it that it would not be static.
# Yes this hurts me.
volcanoes <- read_csv("volcanoes.csv")
visited <- tibble(
  Volcanoes = c("Maungauika / North Head", "Takarunga / Mount Victoria", "Albert Park Volcano", "Maungakiekie / One Tree Hill", "Te Kōpuke / Tītīkōpuke / Mount St John", "Ōhinerau / Mount Hobson", "Ohuiarangi / Pigeon Mountain", "Te Kopua Kai-a-Hiku / Panmure Basin", "Onepoto", "Maungarei / Mount Wellington", "Taurere / Taylors Hill", "Whakamuhu / Saint Heliers / Glover Park", "Maungawhau / Mount Eden"),
  Visited = T, # Redundant column for future filters or whatever idk
  `Date visited` = c("2023-07-11", "2023-07-11", "2023-07-17", "2023-07-22", "2023-07-22", "2023-07-22", "2023-07-30", "2023-08-03", "2023-08-03", "2023-08-27", "2023-08-27", "2023-08-27", "2023-09-01")
)

bigvisited <- inner_join(volcanoes, visited) %>% 
  select(Volcanoes, `Date visited`, Age, `Height (ft)`, Location) %>% 
  mutate(
    `Date visited` = as.Date(`Date visited`)
  )

summaryvisited <- inner_join(volcanoes, visited) %>% 
  summarize(
    Location = paste0("n = ", n(), " / ", nrow(volcanoes)),
    `Height (ft)` = sum(`Height (ft)`, na.rm = T),
    Age = as.integer(mean(Age), na.rm = T),
    Volcanoes = "Mean Age / Total Height"
  )

knitr::kable(
  bind_rows(bigvisited, summaryvisited) %>% 
    arrange(`Date visited`, Volcanoes)
)
```
Last updated 2023-09-02. 

## Code availability

The source code for this project is available at [https://github.com/legallyahc/volcanic](https://github.com/legallyahc/volcanic), with all work done right in this RMarkdown file (`dataset.rmd`) :3. 

## Copyright info

This project is based on information from Wikipedia which was authored by Wikipedia contributors and is licensed under [CC-BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/). 
This project is licensed under the [GNU General Public License v3](https://www.gnu.org/licenses/gpl-3.0.html) 
or later, a compatible license with CC-BY-SA 4.0.

<br>

Originally published on 2023-07-16. Version 1.2.

##### [Back to En cours homepage](../index.html)